---
title: "PA1_template.Rmd"
author: "Premjith Balakrishnan"
date: "August 14, 2015"
output: html_document
---

This work is submitted as part of Peer Assessment 1 requirement within the Reproducible Research course.

Load all the necessary libraries and set the right options

```{r}
library(knitr)
library(data.table)
library(lattice)

opts_chunk$set(echo = TRUE, results = 'hold')

```

Load the activity data and create a data frame with the necessary data elements

```{r}
# Load the data from input file
fit.data <- read.csv("activity.csv", header = TRUE, sep = ",",
                     colClasses = c("numeric", "character", "numeric"))

# Extract the date from the text field into a date element
fit.data$date <- as.POSIXct(fit.data$date, format="%Y-%m-%d")

# Make it into a data frame also including the weekday from date
fit.data <- data.frame(date=fit.data$date, 
                       dayofweek=tolower(weekdays(fit.data$date)), 
                       steps=fit.data$steps, 
                       interval=fit.data$interval)

# Append a daytype column to distinguish between weekday and weekend
fit.data <- cbind(fit.data, 
                  daytype=ifelse(fit.data$dayofweek %in% c("saturday", "sunday"), 
                                 "weekend", 
                                 "weekday"))

# Rearrange the columns
fit.data <- data.frame(date=fit.data$date,
                       interval=fit.data$interval,
                       dayofweek=fit.data$dayofweek,
                       daytype=fit.data$daytype,
                       steps=fit.data$steps)


```
This is the final data frame that has been put together
```{r}

head(fit.data)

```
<b>What is mean total number of steps taken per day?</b>

Computing aggregate statistics to display
```{r}
fit.steps <- aggregate(steps ~ date, data=fit.data, sum, na.rm=TRUE)
```

1. Here is a histogram of the number of steps per day

```{r}
hist(fit.steps$steps, 
     main="Steps per day", 
     breaks=25,
     xlab="Steps", 
     ylab="Frequency", 
     col="red")
```

2. And some descriptive statistics

Mean: 
```{r} 
fit.steps.mean <- mean(fit.steps$steps)
cat("Mean:", fit.steps.mean)
```

Median:
```{r} 
fit.steps.median <- median(fit.steps$steps)
cat("Median:", fit.steps.median)
```

<b>What is the average daily activity pattern?</b>

```{r}
# Compute the average number of steps

fit.avgsteps <- aggregate(steps ~ interval, data=fit.data, mean, na.rm=TRUE)

head(fit.avgsteps)

```

1. Time series plot of the average number of steps taken 

```{r}
plot(fit.avgsteps$interval, 
     fit.avgsteps$steps, 
     type="l", 
     col="blue", 
     xlab="Time Interval", 
     ylab="Average Steps Taken", 
     main="Time series plot of steps taken")

```


2. When was the maximum number of steps taken?

```{r}
# Locate when the maximum steps happened in the time series

fit.avgsteps.maxloc <- which(fit.avgsteps$steps == max(fit.avgsteps$steps))
cat("The max happened at", fit.avgsteps[fit.avgsteps.maxloc,1])

```

<b>Imputing missing values</b>

1. Calculating and reporting missing values

```{r}
# Count the records with NA
fit.data.NA <- sum(ifelse(is.na(fit.data$steps),1,0))

# Report the NA count
cat("Number of records with NA is", fit.data.NA)

```
2. Devising a strategy for filling in missing values


```{r}
# Find the missing indices
fit.data.missing <- is.na(fit.data$steps)

# Create a copy of the original dataset
fit.data.imputed <- fit.data

# Assign the mean value for the steps element in the table for all those indices
# Using a simple approximation for steps per interval from the daily averages
fit.data.imputed[fit.data.missing,"steps"] <- fit.steps.mean / (24*12)

```

3. Create a new data set with the missing data filled in

```{r}

# Since we created a copy in the earlier step 
# it now becomes our new dataset with missing filled in

head(fit.data.imputed)

# Confirming that there are no missing values

cat("Missing values", sum(ifelse(is.na(fit.data.imputed$steps),1,0)))

```
4. Reporting with histogram and descriptive statistics

Computing aggregate statistics to display
```{r}
# Aggregated for each day
fit.steps.imputed <- aggregate(steps ~ date, data=fit.data.imputed, sum, na.rm=TRUE)
```

```{r}
# Draw the histogram
hist(fit.steps.imputed$steps, 
     main="Steps per day", 
     breaks=25,
     xlab="Steps", 
     ylab="Frequency", 
     col="green")

```

Calculate the mean and median

Mean: 
```{r} 
fit.steps.imputed.mean <- mean(fit.steps.imputed$steps)
cat("Mean:", fit.steps.imputed.mean)
```

Median:
```{r} 
fit.steps.imputed.median <- median(fit.steps.imputed$steps)
cat("Median:", fit.steps.imputed.median)
```

Comparison of mean and median after imputing values for missing

```{r}

cat("Mean in the initial dataset:", fit.steps.mean, 
    " but after imputing:", fit.steps.imputed.mean)
cat("\nMedian in the initial dataset:", fit.steps.median, 
    " but after imputing:", fit.steps.imputed.median)

```
After filling in for missing values it can be seen that the mean remains unchanged but the median is now same as the mean. This is expected as the initial mean calculation ignored missing values and the filling in was done with the mean value. However, as for the median, with all the NA now taking value equal to the mean, it looks like the median value is now same as the mean. 


<b>Differences between weekdays and weekends</b>

1. New factor variable for telling "weekday" vs "weekend"

```{r}

head(fit.data.imputed)
cat("The \"daytype\" column provides the info if its a weekday or weekend")

```

2. Make a panel plot containing a time series plot

```{r}
# Compute the time series
fit.avgsteps.daytype <- aggregate(fit.data.imputed$steps,
                                  by=list(fit.data.imputed$daytype,
                                          fit.data.imputed$dayofweek,
                                          fit.data.imputed$interval),
                                  mean)

head(fit.avgsteps.daytype)

# Rename the attributes
names(fit.avgsteps.daytype) <- c("daytype", "weekday", "interval", "mean")

# Plot the data

xyplot(mean ~ interval | daytype, fit.avgsteps.daytype, 
       type="l", 
       lwd=1, 
       xlab="Interval", 
       ylab="Number of steps", 
       layout=c(1,2))

```




